From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-19 paper/paper.Rmd
WIP

## Use trends style to quickly count refs
#csl: trendsjournals.csl
csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

@@

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena.  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.
I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: ecology is complex, and much effort
has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

<!-- Shorten this, more citations, more focus -->

What is noise? Where does it come from?  Simply put, *noise is all that we leave out of our model*. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochastic summary and underlying process more precise:
that is, on understanding the origins of noise [@Coulson2004].

Over the past several decades, research into stochastic modeling
has

@@

With the exception of quantum mechanics (and there only at scales
that have no relevance to ecology), stochasticity is not an inherent
property of nature but rather an emergent property of many events
occurring at a different scale.  Coins and dice obey the physical
laws of motion; only variation in how they are tossed (i.e. something
complex we do not usually model) creates randomness.  The term
"noise" itself reflects this emergent rather than fundamental origin:
the combination of hundreds of voices and sounds each individually
distinct blends together into a background of noise. Statistical
mechanics tells us that temperature is a measure of the average
speed of individual molecules in a gas -- any given molecule will
experience a range of possible speeds but by itself has no notion
of temperature.  Likewise, stochasticity enters our ecological
models because our quantities of interest, such as the density of
a population, is but an average over a lower-level, discrete process
of individual births and deaths. Stochasticity that emerges directly
from discrete individual events that underlie the process described
by our model are considered *intrinsic stochasticity*, commonly
referred to as *demographic stochasticity* in ecological literature,
as it usually arises in the context scaling from individual demographic
events to population level descriptions. An important aspect of
this demographic stochasicity is that it scales with the square
root of the "system size," [@vanKampen2007] so that for very large populations the
resulting fluctuations are relatively small. On the other hand,
*extrinsic stochasticity*, usually referred to as *environmental
stochasticity* in the ecological literature, arises in processes
that are not described by our model, but merely impact the value
of some parameter of our model; for instance, fluctuations in
temperature may drive variation in a death rate, or fluctuations
in rainfall drive variation in carrying capacity.

>>>>

## Use trends style to quickly count refs
csl: trendsjournals.csl
#csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

@@

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.  This shift in how we think about noise
is nonetheless easy to miss in the extensive literature

Origins of noise and motivation for studying stochastic models @Kendall1949 and @Bartlett1960

This trend from nuisance to creator
to informer is of course not the only or the most obvious trend in the 
extensive literature on stochastic modeling over the past few decades, but rather,
an instructive lens in which to draw out a lar

@Bjornstad2001 @Coulson2001 @Coulson2004

I will focus on examples that
illustrate this larger arc in how we are thinking about stochasticity as the role
it plays.

(e.g. see in particular
@Couslon2004, @Ovaskainen2010 on population extinction with 
a focus on environmental and colored noise, @Black2012 on intrinsic / demographic noise,

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]

Excellent reviews and textbook introductions

@Bartlett1960 @Bjornstad2001 @Coulson2001 @Lande2003

The extensive literature on demographic and environmental noise has been
frequently reviewed elsewhere [@Black2012]

@Saether1997

I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: after all, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticty literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; species traits
and the interaction of ecological and evolutionary processes [@Ozgul2009; @Lenormand2009; @Vindesens2015; @Schreiber2015],
not to mention the extensive literature in stochastic evolution and popluation genetics
[e.g. Ewens, @Lewontin1969]

and the role of stochasticity in more complex models, particularly in structured populations (matrix models),

in

has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

What is noise? Where does it come from?  Simply put, noise is all that we leave out of our model. 
Whenever stochasticity is introduced into a model, it
is only as a simplification or summary of the effects of a large
number of other processes, often occurring at different scales.
A central theme of stochastic modeling research over the past
several decades has been an emphasis on making this relationship
between stochasticity and underlying processes more precise:
that is, on the origins of noise [e.g. @Coulson2004]. This literature
has largely divided the origins of stochasticity between 
*intrinsic* factors: where stochasticity is introduced to reflect
the reality that our variables of interest are intriniscally 
statistical averages of lower-level processes, and *extrinsic*
factors: where stochasticity reflects the impact of other variables
we are not explicitly modeling at all [e.g. @Keeling2012].  **Demographic stochasticity**
is the primary (though not the only) example of an intrinsic factor,
which reflects the fact that the continuous variable we often think of a
"population size" or "population density" is an average over a 
lower level process of births and deaths among a finite number of 
individuals [@Melbourne2012].  **Environmental stochasticity** is the primary example 
of an extinsic factor, reflecting the fact that parameters we may
treat as constants in a model in fact depend on variables such as 
temperature or soil moisture that are not explicitly modeled by our 
equations, but simplified into statistical summaries of those processes
[@Ripa2012].

The Gillespie simulation provides a simple and exact algorithm
for simulating demographic stochasticity at an individual level.  
Free from both the approximations and mathematical complexity of more common
formulations using stochastic differential equations (SDEs), 
(or their partial differential equation counterparts the Fokker-Planck equation),
the Gillespie algorithm offers a simple and powerful approach to simulating
stochastic phenomena that remains underutilized.

The mathematical proofs @Kurtz1970 & @Kurtz1971, popularized by
@vanKampen's intuitive "system size" expansion, provide a far more 
explicit framework to interpret demographic stochasticity than the
better known diffusion approximation [@Nisbet1982 @Lande2003; 
see @Gardiner2009].

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality.  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinisic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises
only subsequently, when we change scales from a description at the individual
level to one at the population level.

@Keeling2012; @Melbourne2012; @Ripa2012

@@
